Metabolites reflect variability introduced by mesoscale eddies in the North Pacific Subtropical Gyre

### 2024-10-30 manuscript/manuscript.Rmd
Responding to reviewer #1 concerns
Needs additional feedback from coauthors

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, eval = TRUE)
library(tidyverse)
library(RaMS)
library(vegan)
library(cowplot)
library(viridis)
library(scales)
library(ggnewscale)
library(egg)
library(broom)
library(lme4)
library(openxlsx)
library(readxl)
library(ggh4x)
library(ggtext)
library(gridExtra)
library(openxlsx)
library(arrow)

@@

The strong SLA signal detected among the 175 meter samples was partially surprising to us given our expectation about the strongest effect at the DCM where eddy effects lift large concentrations of nutrients above the 1% light level. Instead, we found more compounds overall to be significantly different across the eddy transect in the 175 meter samples than we did at the DCM (α=0.05, 43 compounds at 175 meters vs 22 at the DCM). This was likely driven by the enhancement of organic matter and heterotrophic picoplankton below the DCM as seen in @Barone2019, with the enrichment of nucleobases in the 175 meter samples additionally hint at increases in bacterial biomass given that nucleobases tend to be highly abundant in bacterial cultures [@Heal2021]. The abundance of a mass feature putatively identified as acetylcholine in the 175 meter cyclone samples followed the same general trend as arsenobetaine, both enriched at depth and in the anticyclone, and may be another compound that results from heterotrophic degradation of phytoplankton metabolites [@Durham2022]. These results raise important questions about the depth at which the community is isolated from SLA effects, with additional data from @Barone2022 and @Gleich2024 indicating that even at 250 meters eddy effects are still discernable.

Differences between the two cruises could be due to differences in eddy age and/or the different period of the year when sampling took place during the two cruises (March/April in 2018 instead of the June/July sampling in 2017) resulting in shifts in community composition and response. @Barone2022 documents the 2017 cruise community composition, reporting abundant Prymnesiophyceae at both 15 meters and the DCM, with other significant contributions from dinoflagellates, diatoms, and green algae. In the 2018 cruise both haptophytes and dinoflagellates again dominated the eukaryotic fraction, though surprisingly little was contributed by diatoms [@Gleich2024]. @Dugenne2023 and @Harke2021 also highlight uncontrolled differences between the two sets of eddies based on their origin and age.

Missing links between the environment with its associated community and the composition of particulate matter make it difficult to estimate this important control on the marine carbon cycle. This is especially true for dynamic and short-lived environments such as eddies. Metabolites offer a way to bridge this gap in knowledge but the relative impact of eddies on their contribution and fate was not previously estimated. In this work, we have shown the sensitivity of the metabolome to various environmental and community factors and contrasted it to the well-established effect of depth to constrain the importance of including eddy effects in models of elemental cycling.

@@

This study reports how changes in the biogeochemistry and community composition due to mesoscale eddies affect the metabolome and thereby the composition of particulate organic matter. A transect across adjacent mesoscale eddies of opposite polarity showed that many metabolites track closely with metrics of overall biomass, with eddy effects stronger at 175 meters than at 15 meters or the deep chlorophyll maximum (DCM). High-resolution depth sampling of the DCM at the center of each eddy elucidated several known and unknown biomarkers likely corresponding to the previously documented increase in eukaryotic phytoplankton in the cyclone. Metabolite clusters aligned well with expected trends in abundance with depth and between the two eddies, with convergence below the DCM towards a general deep-sea metabolite signal for many metabolites. By contrasting these effects with a follow-up analysis in the centers of a separate pair of mesoscale eddies of opposite polarity the following year, we learned that the impact of eddies may be influenced by their origin, age, and period of the year. Finally, the metabolites with the strongest responses to eddy effects were unknown, and a relatively smaller proportion of known metabolites in the anticyclone demonstrates the utility of untargeted metabolomics in exploring environmental variation and the large amount of information that is missed by only investigating known compounds.

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, eval = FALSE)
library(tidyverse)
library(RaMS)
library(vegan)
library(cowplot)
library(viridis)
library(scales)
library(ggnewscale)
library(egg)
library(broom)
library(lme4)
library(openxlsx)
library(readxl)
library(ggh4x)
library(ggtext)
library(gridExtra)
library(openxlsx)
library(arrow)

@@

The strong SLA signal detected among the 175 meter samples was partially surprising to us given our expectation about the strongest effect at the DCM where eddy effects lift large concentrations of nutrients above the 1% light level. Instead, we found more compounds overall to be significantly different across the eddy transect in the 175 meter samples than we did at the DCM (α=0.05, 43 compounds at 175 meters vs 22 at the DCM). This was likely driven by the enhancement of organic matter and heterotrophic picoplankton below the DCM as seen in @Barone2019, with the enrichment of nucleobases in the 175 meter samples additionally hinting at increases in bacterial biomass given that nucleobases tend to be highly abundant in bacterial cultures [@Heal2021]. The abundance of a mass feature putatively identified as acetylcholine in the 175 meter cyclone samples followed the same general trend as arsenobetaine, both enriched at depth and in the anticyclone, and may be another compound that results from heterotrophic degradation of phytoplankton metabolites [@Durham2022]. These results raise important questions about the depth at which the community is isolated from SLA effects, with additional data from @Barone2022 and @Gleich2024 indicating that even at 250 meters eddy effects are still discernable.

## Metabolites link changes in biogeochemistry to community biology

Missing links between the environment with its associated community and the composition of particulate matter make it difficult to estimate important controls on the marine carbon cycle. This is especially true for dynamic and short-lived environments such as eddies. Metabolites offer a way to bridge this gap in knowledge by directly measuring the units of exchange between organisms and their environment.

Here, we saw how changes in the density surface driven by physics during this process led to the coupling of physics, chemistry, and biology. The enrichment of eukaryotic organisms at the DCM of the cyclone mirrors other ocean processes where upwelling of deep, nutrient-rich water leads to increases in nitrogen-rich metabolites representative of a shift towards organisms such as diatoms and dinoflagellates that undergo bloom dynamics. @Barone2022 established a "chemical wake" hypothesis in which cyclone formation introduces a transient phytoplankton community with net photosynthetic production that induces export, a conceptual model that can be extended to include many of the metabolites named here, as some metabolites are more bioavailable than others and the "flavor" of this chemical wake will change over time.

Almost all large phytoplankton classes in the 2017 eddy pair were enriched in the cyclone DCM relative to the anticyclone (particularly prymnesiophytes, dinoflagellates, diatoms, and green algae), while *Prochlorococcus* was significantly less abundant and *Synechococcus* was approximately equal [@Barone2022]. Thus, the metabolite data presented here does not help distinguish between large phytoplankton classes but instead provides additional cues about the shift from a metabolome dominated by cyanobacteria producers into one dominated by eukaryotic organisms. This shift occurs anywhere

A shift from targeted to untargeted metabolites may also in part explain why the increased biomass in the cyclone did not result in large increases in export [@Barone2022; @Bent2024]. Our list of known compounds covers a large fraction of the small molecules common to most organisms, indicating that unknown molecules are more likely to be rare and therefore more difficult to degrade. If the increased biomass in the cyclone is dominantly in the form of highly bioavailable compounds, it is not surprising that much of it can be easily degraded. This hypothesis is supported by @Gleich2024, who found that heterotrophy-associated transcripts were more abundant in the 2018 cruise cyclone than the anticyclone.

Both cruises occurred in weakening cyclones and stable anticyclones, limiting the degree to which the results presented here can be extrapolated to the full life cycle of an eddy [@Dugenne2023]. Differences between the 2017 and 2018 cruises in both biogeochemistry and community biology additionally make it difficult to generalize the links between these two factors via metabolites and establish the demand for additional characterization of the metabolome in the North Pacific Subtropical Gyre.

@@

This study reports how changes in the biogeochemistry and community composition due to mesoscale eddies affect the metabolome and thereby the composition of particulate organic matter. A transect across adjacent mesoscale eddies of opposite polarity showed that many metabolites track closely with metrics of overall biomass, with eddy effects stronger at 175 meters than at 15 meters or the deep chlorophyll maximum (DCM). High-resolution depth sampling of the DCM at the center of each eddy elucidated several known and unknown biomarkers likely corresponding to the previously documented increase in eukaryotic phytoplankton in the cyclone. Metabolite clusters aligned well with expected trends in abundance with depth and between the two eddies, with convergence below the DCM towards a general deep-sea metabolite signal for many metabolites. By contrasting these effects with a follow-up analysis in the centers of a separate pair of mesoscale eddies of opposite polarity the following year, we learned that the impact of eddies may be influenced by their origin, age, and period of the year. Finally, the metabolites with the strongest responses to eddy effects were unknown, and a relatively smaller proportion of known metabolites in the anticyclone demonstrates the utility of untargeted metabolomics in exploring environmental variation and the large amount of information that is missed by only investigating known compounds. In this work, we have shown the sensitivity of the metabolome to various environmental and community factors and constrained the importance of eddy effects in models of elemental cycling.
